---
title: "Decomposition tables and figures for Stovitz, Wolfson, Shrier et al."
author: "J. Wolfson"
date: "Last updated: October 28, 2016"
output: html_document
params:
  code.folder: "C:/Users/Julian/Google Drive/git/ACLS-decomposition/"
  dat.folder: "C:/Users/Julian/Google Drive/git/ACLS-decomposition/data/"
  dat.filename: "acls-fake.csv"
  savePlots: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(survival)
library(boot)
library(knitr)

#dat.base <- paste0(getwd(), "/", params$dat.folder)
dat.base <- params$dat.folder
code.base <- params$code.folder
#code.base <- paste0(getwd(), "/", params$code.folder)

source(paste0(code.base, "fourway-decomp-functions.R"))

pretty.pvalue <- function(p, sdigits = 2, prefix = "p = ") { ## Utility function for displaying p-values
  return( ifelse(p<0.001,"p < 0.001",sprintf("%s%s",prefix, as.character(signif(p,sdigits))) ) )
}

####### Read in the data ###########
acls <- read.csv(paste0(dat.base, params$dat.filename))
acls$PA <- acls$TOTALmet
acls$CRF <- acls$testmin
acls$PA.ir <- as.integer(acls$PA>0) ## Insufficient or recommended
acls$PA.r <- as.integer(acls$PA >= 500)
acls$CRF.mh <- as.integer( acls$CRF >= stats::quantile(acls$CRF, 0.2) ) ## Moderate or high
acls$CRF.h <- as.integer( acls$CRF >= stats::quantile(acls$CRF, 0.6) ) ## High
acls$BMI.h <- as.integer( acls$BMI >= median(acls$BMI) )

upstream.confounders <- c("AGE","female","examyy","smoking","fCVD", "BMI.h")

attach(acls)

surv.obj <- Surv(followyear, deceased)

fitAFT <- function(trt, confounders, transf = function(x) exp(x), dat = acls, dist = "loglogistic") {
  surv.obj <- with(dat, Surv(followyear, deceased))
  fmla <- as.formula( sprintf("surv.obj ~ %s + %s", trt, paste(confounders, collapse = " + ")) )
  S <- summary( SR <- survreg( fmla, data=dat, dist = dist ) )
  summ.table <- cbind( transf(S$coefficients), t( apply( transf(confint(SR)),1, rev ) ), S$table[-nrow(S$table),4] )
  colnames(summ.table)[4] <- "p"
  return( summ.table[-1,] )
}

exposure.effect <- function(T, CI = TRUE) { 
  if(CI) {
    CI.vals <- sort( c( T[1,2], T[1,3] ))
    
    rbind( sprintf("%.2f", T[1,1]), 
          sprintf("(%.2f, %.2f)", CI.vals[1], CI.vals[2] )  )    
  } else {
    sprintf("%.2f (%s)", T[1,1], pretty.pvalue(T[1,4])  )    
  }
}
```

```{r include=FALSE, cache=TRUE}
### Run the mediation analysis and cache the results ###
acls.male <- subset(acls, female == 0)
acls.female <- subset(acls, female == 1)

upstream.confounders <- c("AGE","female","examyy","smoking","fCVD", "BMI")
upstream.confounders.nosex <- c("AGE","examyy","smoking","fCVD", "BMI")
mediators.PA <- c("CRF.mh", "abnHTN", "abnDM", "abnTC")
mediators.CRF <- c( "PA.r", "abnHTN", "abnDM", "abnTC")

expit <- function(x) exp(x)/(1+exp(x))

R <- 100

### Results for PA as the exposure
mediators <- mediators.PA
trt <- "PA.r"

PA.results <- lapply( mediators, function(mediator) {
    boot(acls, fourway.decomp, trt = trt, mediator = mediator, confounders = upstream.confounders, R = R) })
gc()

PA.results.m <- lapply( mediators, function(mediator) {
  boot(acls.male, fourway.decomp, trt = trt, mediator = mediator, confounders = upstream.confounders.nosex, R = R) })
gc()

PA.results.f <- lapply( mediators, function(mediator) {
  boot(acls.female, fourway.decomp, trt = trt, mediator = mediator, confounders = upstream.confounders.nosex, R = R) })
gc()

acls$age.h <- as.integer(acls$AGE > median(acls$AGE))


### Results for CRF as the exposure
mediators <- mediators.CRF
trt <- "CRF.mh"

CRF.results <- lapply( mediators, function(mediator) {
  boot(acls, fourway.decomp, trt = trt, mediator = mediator, confounders = upstream.confounders, R = R) })
gc()

CRF.results.m <- lapply( mediators, function(mediator) {
  boot(acls.male, fourway.decomp, trt = trt, mediator = mediator, confounders = upstream.confounders.nosex, R = R) })
gc()

CRF.results.f <- lapply( mediators, function(mediator) {
  boot(acls.female, fourway.decomp, trt = trt, mediator = mediator, confounders = upstream.confounders.nosex, R = R) })
gc()

```

## Exposure: PA (Recommended vs. Insufficient/Inactive)

```{r include=FALSE}
# Adjusted associations
a11 <- exposure.effect( fitAFT("PA.r", confounders = upstream.confounders) )
a12 <- exposure.effect( fitAFT("PA.r", confounders = c( upstream.confounders, "CRF.mh") ) )
a13 <- exposure.effect( fitAFT("PA.r", confounders = c( upstream.confounders, "abnHTN") ) )
a14 <- exposure.effect( fitAFT("PA.r", confounders = c( upstream.confounders, "abnDM") ) )
a15 <- exposure.effect( fitAFT("PA.r", confounders = c( upstream.confounders, "abnTC") ) )

mediators.PA <- c("CRF.mh", "abnHTN", "abnDM", "abnTC")

PA.tables <- lapply(list( PA.results ), table.fourway, mediators = mediators.PA, transf = function(x) exp(-x))

```

```{r echo=FALSE}
na.mat <- matrix(data="N/A", nrow = 2, ncol = 3)
df <- cbind( rbind( a11,a12,a13,a14,a15 ), rbind(na.mat, PA.tables[[1]][,c(1,13,14)]) )
colnames(df) <- c("Adjusted Association", "Total Effect", "% Mediated", "% Interaction")
rownames(df) <- c("Upstream Only", "", "+ CRF", "", "+ HTN", "", "+ Diabetes", "", "+ HighChol", "")
kable( df, 
       row.names = TRUE)
```

```{r include = FALSE}
getPct <- function(str, direction.word = TRUE) {
  pct <- as.numeric(str)
  if(direction.word) {
    sprintf("%.0f%% %s", abs( round(100 * (pct - 1)) ), ifelse(sign(pct)==1, "increase", "decrease") )
  } else {
    sprintf("%.0f%%", round(100 * (pct - 1)) )
  }
}

getCI <- function(str) {
  strip.l <- gsub( "\\(", "", str)
  strip.r <- gsub("\\)", "", strip.l)
  strip.c <- gsub(",", "", strip.r)
  get.vals <- as.numeric(strsplit(strip.c, " ")[[1]])
  pcts <- sort(get.vals)
  
  sprintf("%.0f%%,  %.0f%%", round(100 * (pcts[1] - 1)), round(100 * (pcts[2] - 1)) )
}
```

As seen in Table 1, after adjusting for all upstream variables, recommended PA (vs insufficient/inactive) was associated with approximately a `r getPct(a11[1,1])` in the median time to death, (95% CI: `r getCI(a11[2,1])`). Basic adjustment for potential mediators resulted in a smaller effect of PA when the mediator was CRF (i.e., `r getPct(a12[1,1])`, 95% CI: `r getCI(a12[2,1])`), but an unchanged effect of PA with the other mediators, (HTN, DM, and High Cholesterol). Decomposition analysis found that `r df[3, "% Mediated"]` (95% CI: `r df[4, "% Mediated"]`) of the effect of PA was mediated by CRF. A small percentage of the effect of PA was mediated by HTN and Diabetes. There was no evidence of statistically significant interaction between PA and CRF, HTN, DM, or High Cholesterol. 

```{r fig.width=11, fig.height=6, warning=FALSE}
library(ggplot2)

exposure <- "Physical Activity"
mediator.names <- c("Cardiorespiratory Fitness", "Hypertension", "Diabetes", "Hypercholestrolemia")
g.PA <- gg.fourway( PA.results, exposure, mediator.names, custom.ylim = c(0.88, 1.25) )

multiplot(plotlist = g.PA, cols = 2)
```

## Exposure: CRF (Fit vs. Unfit)

```{r include = FALSE}
b11 <- exposure.effect( fitAFT("CRF.mh", confounders = upstream.confounders) )
b12 <- exposure.effect( fitAFT("CRF.mh", confounders = c( upstream.confounders, "PA.r") ) )
b13 <- exposure.effect( fitAFT("CRF.mh", confounders = c( upstream.confounders, "abnHTN") ) )
b14 <- exposure.effect( fitAFT("CRF.mh", confounders = c( upstream.confounders, "abnDM") ) )
b15 <- exposure.effect( fitAFT("CRF.mh", confounders = c( upstream.confounders, "abnTC") ) )

mediators.CRF <- c( "PA.r", "abnHTN", "abnDM", "abnTC")

CRF.tables <- lapply(list( CRF.results ), table.fourway, mediators = mediators.CRF, transf = function(x) exp(-x))

```

```{r}
na.mat <- matrix(data="N/A", nrow = 2, ncol = 3)
df <- cbind( rbind( b11,b12,b13,b14,b15 ), rbind(na.mat, CRF.tables[[1]][,c(1,13,14)]) )
colnames(df) <- c("Adjusted Association", "Total Effect", "% Mediated", "% Interaction")
rownames(df) <- c("Upstream Only", "", "+ PA", "", "+ HTN", "", "+ Diabetes", "", "+ HighChol", "")
kable( df, 
       row.names = TRUE)
```

Table 2 describes the effect of CRF on mortality. After adjusting for all upstream variables, those who were "fit" (vs unfit) had approximately a `r getPct(b11[1,1])` in the median time to death, (95% CI: `r getCI(b11[2,1])`). Basic adjustment for potential mediators had minimal to no effect on the rate. Decomposition analysis found that a small amount of the effect of CRF appeared to mediated by HTN, DM and High Cholesterol. There was no significant mediation of the effect of CRF by PA. 

```{r  fig.width=11, fig.height=6}
exposure <- "Cardiorespiratory Fitness"
mediator.names <- c("Physical Activity", "Hypertension", "Diabetes", "Hypercholesterolemia")
g.CRF <- gg.fourway( CRF.results, exposure, mediator.names, custom.ylim = c(0.94, 1.55) )

multiplot(plotlist = g.CRF, cols = 2)
```